Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2019-09-27 manuscript/main.tex
Use better metric names, flesh out results for natural populations
Consistently uses "HI cross-immunity" instead of "HI phenotypes" and "DMS
mutational effect" instead of "DMS phenotypes".

Here, we present a model of seasonal influenza evolution inspired by the exponential growth model of \cite{Luksza:2014hj}.
As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying an exponential growth factor to each virus sample scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model.
With this approach, we avoid the intrinsic variability of ancestral sequence reconstruction associated with previous clade-based models.
However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.

@@

The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions at epitopes that enable escape from adaptive immunity and purifying selection on domains required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under purifying and positive selection for 40 years (Methods).
These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
We fit models to these simulated populations using all sequence-only fitness metrics.

We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while the non-epitope mutations metric would be assigned a negative coefficient.
We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
Correspondingly, we expected that a combined model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics.
In addition to these four estimates of viral fitness, we tested models based on both the true fitness of each sample as measured by the simulator and a naive model under which the exponential growth factor is set to one and populations do not change composition over the one year forecasting period.
The distances estimated under the naive model represent the average distance between the current and future timepoints.

@@

The average distance between yearly populations, as measured by the naive model, was 8.97 +/- 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.17 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection} and Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
With the exception of epitope cross-immunity, all biologically-informed models also outperformed the naive model.
LBI was the best of these models, reducing the distance between populations by 1.41 amino acids on average.
Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
The non-epitope mutations metric received a consistently negative coefficient and also outperformed the naive model with an average improvement of 0.71 amino acids.
Surprisingly, the combined model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions was dominated by purifying selection rather than by positive selection at epitope sites.

We hypothesized that a combined model of mutually beneficial metrics could better approximate the true fitness of simulated viruses.
To this end, we fit an additional combined model including both LBI and non-epitope mutations.
This combined model outperformed all individual metrics, reducing the distance between populations by 1.76 amino acids and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
Interestingly, we found that the coefficients for LBI and non-epitope mutations remained relatively stable across all validation timepoints.
These results support our hypothesis that multiple complementary metrics can produce more accurate models.

Finally, we sought to validate and test the best performing model by two metrics relevant for practical influenza forecasting and vaccine design efforts.
First, we measured the ability of the best model to accurately estimate clade dynamics by correlating estimated and observed clade growth rates.
Under these simulated conditions, our model satisfactorily recapitulated clade dynamics with a growth rate correlation of $R = 0.71$ $(p < 0.001)$ and overall accuracies for clade growth and decline predictions of 66\% and 61\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
Next, we tested how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
The estimated best strain was in the top 4th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 82\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).

@@

Next, we trained and validated models for individual fitness predictors using 23 years of natural A/H3N2 populations spanning from October 1992 to October 2015.
We held out samples collected between October 2015 and October 2019 for model testing.
In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes from HI and DMS assays.
We hypothesized that both HI and DMS phenotype metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope cross-immunity and HI phenotypes would be individually more predictive than non-epitope mutations or DMS phenotypes.
Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.

@@

The average distance per year between natural populations was 6.50 +/- 1.42 amino acids or 72\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS phenotypes metrics (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
Surprisingly, the best antigenic fitness metric of HI phenotypes performed only slightly better the best functional constraint metric of non-epitope mutations.
Indeed, epitope cross-immunity only outperformed the naive model at one of 19 timepoints (5\%), while HI phenotypes outperformed the naive model at 15 (79\%) timepoints (Table~\ref{table_natural_model_selection}).
Epitope cross-immunity was also the only metric whose coefficient started at a positive value and transitioned to a negative value through the validation period.
This change in coefficient suggests that positive selection may have weakened at these epitope sites over time.
In contrast, HI phenotypes maintained a positive coefficient across most timepoints.
The HI phenotypes metric may benefit from being able to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.

Non-epitope mutations also outperformed the DMS phenotypes metric, reducing the distance to the future by 0.40 amino acids on average compared to 0.19 amino acids, respectively.
In contrast to \cite{Luksza:2014hj}, where the coefficient of the non-epitope mutations metric was hardcoded to -0.5, our model learned a consistently stronger coefficient of -1.45 +/- 0.24.
Notably, the DMS metric only performed noticeably better than the naive model at the timepoint immediately preceding when the background strain of the DMS experiments, A/Perth/16/2009, was sampled.
This result is consistent with the DMS model overfitting to the evolutionary history of the DMS background strain used to perform the DMS experiments.
Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Fig.~\ref{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}).
These results suggest that any mutation at a non-epitope site is deleterious for globally circulating strains, despite the more fine-grained fitness constraints of within-host infections.

LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
However, LBI only outperformed the naive model 63\% of the time (Table~\ref{table_natural_model_selection}).
These results suggest that when LBI forecasts correctly, it does so better than other metrics, but that it is prone to overfitting.
Delta frequency did not surpass LBI in average estimated distance to the future and performed worse than both the HI phenotypes and non-epitope mutation models.
While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.

@@

To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
Specifically, we fit models based on HI phenotypes and non-epitope mutations, LBI and non-epitope mutations, and all three of these metrics together.
We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.

@@

As expected, all of these combined models performed better than any of their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
The model with all three metrics performed the best, reducing the distance to the future by 1.21 amino acids on average.
However, the composite of LBI and non-epitope mutations performed nearly as well with an average reduced distance of 1.14.
Surprisingly, the combination of HI phenotypes and non-epitope mutations did not perform better than LBI by itself, despite the stability of the coefficients for both metrics across all timepoints.

@@

As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
We find that the combined model of LBI and non-epitope mutations accurately captures clade decline and modestly reflects observed clade growth (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
  \caption{
  Validation of best model for natural populations of A/H3N2 viruses, the combined model of LBI and non-epitope mutations.
  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
  }
  \label{fig:validation_of_best_model_for_natural_populations}
  \end{center}
\end{figure*}

>>>>

Here, we present a model of seasonal influenza evolution inspired by the exponential growth model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model (Fig.~\ref{fig:model}).
With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.

@@

The long-term evolution of influenza A/H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \cite{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of A/H3N2-like viruses under positive and purifying selection for 40 years (Methods).
These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural A/H3N2 HA (Supplemental Figure \ref{sup_fig:simulated_h3n2_ha_phylogeny}).
We fit models to these simulated populations using all sequence-only fitness metrics.

We hypothesized that fitness metrics associated with viral success such as epitope cross-immunity, LBI, and delta frequency would be assigned positive coefficients, while metrics associated with fitness penalties, like non-epitope mutations, would receive negative coefficients.
We reasoned that both LBI and delta frequency would individually outperform the mechanistic metrics as both of these growth metrics estimate recent clade success regardless of the mechanistic basis for that success.
Correspondingly, we expected that a composite model of epitope cross-immunity and non-epitope mutations would perform as well as or better than the growth metrics.
In addition to these four estimates of viral fitness, we tested models based on both the true fitness of each sample as measured by the simulator and a naive model under which the exponential growth factor is set to one and populations do not change composition over the one year forecasting period.
The distances estimated under the naive model represent the average distance between the current and future timepoints.

@@

The average distance between yearly populations, as measured by the naive model, was 8.97 +/- 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.17 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection} and Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
With the exception of epitope cross-immunity, all biologically-informed models also outperformed the naive model.
LBI was the best of these models, reducing the distance between populations by 1.41 amino acids on average.
Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
The non-epitope mutations metric received a consistently negative coefficient with an average improvement of 0.71 amino acids.
Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions was dominated by purifying selection rather than by positive selection at epitope sites.

We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses.
To this end, we fit an additional model including both LBI and non-epitope mutations.
This composite model outperformed all individual metrics, reducing the distance between populations by 1.76 amino acids and outperforming the naive model as often as the true fitness metric (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
Interestingly, we found that the coefficients for LBI and non-epitope mutations remained relatively stable across all validation timepoints.
These results support our hypothesis that multiple complementary metrics can produce more accurate models.

Finally, we sought to validate and test the best performing model by two metrics relevant for practical influenza forecasting and vaccine design efforts.
First, we measured the ability of the true fitness model to accurately estimate clade dynamics by correlating estimated and observed clade growth rates.
Under these simulated conditions, our model satisfactorily recapitulated clade dynamics with a growth rate correlation of $R = 0.71 (p < 0.001)$ and overall accuracies for clade growth and decline predictions of 66\% and 61\%, respectively (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
Next, we tested how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
The estimated best strain was in the top 4th percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 82\% of timepoints (Supplemental Fig.~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.

@@

Next, we trained and validated models for individual fitness predictors using 23 years of natural A/H3N2 populations spanning from October 1992 to October 2015.
We held out samples collected between October 2015 and October 2019 for model testing.
In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
As antigenic drift is generally considered to be the primary evolutionary pressure on natural A/H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than non-epitope mutations or DMS mutational effects.
Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.

@@

The average distance per year between natural populations was 6.50 +/- 1.42 amino acids or 72\% of the distance between yearly simulated populations (Supplemental Fig.~\ref{sup_fig:distance_of_natural_populations_between_timepoints}).
Biologically-informed metrics generally performed better than the naive model for natural populations with the exceptions of the epitope cross-immunity and DMS mutational effects (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
Surprisingly, the best antigenic fitness metric of HI cross-immunity performed only slightly better the best functional constraint metric of non-epitope mutations.
Indeed, epitope cross-immunity only outperformed the naive model at one of 19 timepoints (5\%), while HI cross-immunity outperformed the naive model at 15 (79\%) timepoints (Table~\ref{table_natural_model_selection}).
Epitope cross-immunity was also the only metric whose coefficient started at a positive value and transitioned to a negative value through the validation period.
This change in coefficient suggests that positive selection may have weakened at these epitope sites over time.
In contrast, HI cross-immunity maintained a positive coefficient across most timepoints.
The HI cross-immunity may benefit from being able to constantly update its antigenic model at each timepoint with recent experimental phenotypes, while the epitope cross-immunity metric is forced to give a constant weight to the same 49 sites throughout time.

Non-epitope mutations also outperformed the DMS mutational effects, reducing the distance to the future by 0.40 amino acids on average compared to 0.19 amino acids, respectively.
In contrast to the original {\L}uksza and L\"assig \cite{Luksza:2014hj} model, where the coefficient of the non-epitope mutations metric was fixed at -0.5, our model learned a consistently stronger coefficient of -1.45 +/- 0.24.
Notably, DMS mutational effects only performed noticeably better than the naive model at the timepoint immediately preceding when the background strain of the DMS experiments, A/Perth/16/2009, was sampled.
This result is consistent with the DMS model overfitting to the evolutionary history of the background strain used to perform the DMS experiments.
Alternate implementations of less background-dependent DMS metrics never performed better than the non-epitope mutations metric (Supplemental Fig.~\ref{sup_fig:unadjusted_DMS_model_accuracy_and_coefficients_for_natural_populations}).
Thus, we find that a simple model where any mutation at non-epitope sites is deleterious is more predictive of global viral success than a more comprehensive model based on measured mutational effects.

LBI was the best individual fitness metric by average distance to the future (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
However, LBI only outperformed the naive model 63\% of the time (Table~\ref{table_natural_model_selection}).
These results suggest that when LBI forecasts correctly, it does so better than other metrics, but that it is prone to overfitting.
Delta frequency did not surpass LBI in average estimated distance to the future and performed worse than both the HI cross-immunity and non-epitope mutation models.
While delta frequency should, in principle, measure the same aspect of viral fitness as LBI, these results clearly show that the current implementations of these metrics represent qualitatively different fitness components.

@@

To test whether composite models could outperform individual fitness metrics for natural populations, we fit models based on combinations of best individual metrics representing antigenic drift, functional constraint, and clade growth.
Specifically, we fit models based on HI cross-immunity and non-epitope mutations, LBI and non-epitope mutations, and all three of these metrics together.
We anticipated that if these metrics all represented distinct, mutually beneficial components of viral fitness, these composite models should perform better than individual models with consistent coefficients for each metric.

@@

As expected, all of these composite models performed better than any of their corresponding individual models (Table~\ref{table_natural_model_selection} and Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
The model with all three metrics performed the best, reducing the distance to the future by 1.21 amino acids on average.
This 19\% reduction of distance to the future relative to the naive model was nearly equal to the 20\% relative reduction of the best composite model for simulated populations.
However, the composite of LBI and non-epitope mutations performed nearly as well with an average reduced distance of 1.14.
Both of these models were more accurate than the naive model for 79\% of validation timepoints.
Although the composite of HI cross-immunity and non-epitope mutations outperformed the naive model more often than LBI, this composite model did not reduce the estimated distance to the future better on average than the individual LBI metric.
Indeed, we observed that the contribution of HI cross-immunity to the best composite model declined after October 2010 when that metric's coefficient converged to zero (Fig.~\ref{fig:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations}).
These results reinforce the historical importance of HI assays in measuring viral fitness, while also indicating that these assays are no longer as predictive as sequence-only metrics.

@@

As with the simulated populations, we validated the performance of the best model for natural populations using estimated and observed clade growth rates and the ranking of estimated best strains compared to the observed closest strains to future populations.
We find that the composite model of LBI and non-epitope mutations accurately captures clade decline and modestly reflects observed clade growth (Fig.~\ref{fig:validation_of_best_model_for_natural_populations}A).

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation-of-best-model-for-natural-populations.png}
  \caption{
  Validation of best model for natural populations of A/H3N2 viruses, the composite model of LBI and non-epitope mutations.
  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
  }
  \label{fig:validation_of_best_model_for_natural_populations}
  \end{center}
\end{figure*}
